---
title: "Free-living accelerometry with PhysioWearable"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Free-living accelerometry with PhysioWearable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

PhysioWearable analyses raw tri-axial accelerometry following the GGIR
methodology. Its analysis layer is **device-agnostic**: once a recording is a
matrix of acceleration in *g* units, the same functions apply whether the data
came from a GENEActiv, ActiGraph or Axivity device, or from a consumer wearable
export. This vignette uses only synthetic data, so it runs offline and is fully
reproducible --- no device file or API is required.

```{r setup}
library(PhysioWearable)
set.seed(1)
```

## From raw acceleration to ENMO

The core metric is ENMO --- the Euclidean norm of the acceleration vector minus
one gravity, with negatives truncated to zero. A perfectly still sensor (1 *g*)
gives ENMO 0; movement raises it. Here is a two-minute synthetic snippet at
30 Hz: one minute still, then one minute of 1.8 Hz "walking".

```{r raw}
fs  <- 30
t1  <- seq_len(60 * fs) / fs
n1  <- length(t1)
still <- cbind(x = rnorm(n1, 0, 0.01), y = rnorm(n1, 0, 0.01),
               z = 1 + rnorm(n1, 0, 0.01))         # ~1 g on z, sensor still
walk  <- cbind(x = 0.3 * sin(2 * pi * 1.8 * t1), y = 0,
               z = 1 + 0.3 * cos(2 * pi * 1.8 * t1))  # 1.8 Hz movement
accel <- rbind(still, walk)

enmo_epoch <- computeENMO(accel, sampling_rate = fs, epoch_sec = 5, unit = "mg")
round(enmo_epoch)            # ~0 mg while still, higher during "walking"
```

## Step detection and intensity

`detectSteps()` peak-counts the dynamic acceleration; `classifyBouts()` turns
per-epoch ENMO into intensity categories using published mg cut-points.

```{r steps}
detectSteps(accel, sampling_rate = fs)$n_steps

paIntensityThresholds()                 # light / moderate / vigorous, mg
bouts <- classifyBouts(enmo_epoch, epoch_sec = 5, enmo_unit = "mg",
                       min_bout_min = 0)
bouts$minutes
```

## A day-level free-living summary

Day- and person-level metrics are built on per-epoch ENMO. Below is a synthetic
14-hour day in 30-second epochs (mostly sedentary with a one-hour brisk block);
`summarizeFreeLiving()` returns the metrics free-living studies report ---
intensity time-use, the Rowlands intensity gradient, MX metrics, activity
fragmentation and WHO guideline attainment.

```{r day}
ep  <- 30
day <- c(rep(5,   120 * 10),   # 10 h mostly sedentary (5 mg)
         rep(60,  120 * 2),    # 2 h light (60 mg)
         rep(160, 120 * 1),    # 1 h brisk / moderate (160 mg)
         rep(5,   120 * 1))    # 1 h sedentary
summary <- summarizeFreeLiving(day, epoch_sec = ep, enmo_unit = "mg")
summary$aggregate[c("sedentary_min", "light_min", "mvpa_min",
                    "intensity_gradient")]
summary$aggregate$meets_guideline
```

`intensityGradient()` and `mxMetrics()` are also available on their own:

```{r metrics}
round(intensityGradient(day, epoch_sec = ep)$gradient, 2)
mxMetrics(day, epoch_sec = ep, minutes = c(30, 60))
```

## Where next

The free-living metrics link to the ICF Activities and Participation performance
qualifier via `freeLivingICF()` and `adlToICF()`. Raw device exports are read
with `readAccelCSV()` into a PhysioExperiment object; the recording can be
gravity-calibrated with `autoCalibrateAccel()` and screened for non-wear with
`detectNonWear()` before the analysis above.
